Competing Constructions for German Measure Noun Phrases: from Usage Data to Experiment

### 2017-04-09 Latex/article.tex
Minora.

As for the first question, this is not quite the place to discuss it fully.
The basic distinction is a philosophical one and related to the concepts of \textit{direct} and \textit{inverse probability} (\eg\ \citealp{Senn2011}).
Frequentists assume that models and parameters are fixed, for example a model specifying that a coin is fair.
We can then calculate for observed data (for example a measurement of $3$ heads in an experiment with $10$ tosses) how often such a result or a more extreme result would occur if the model were true and we repeated the experiment arbitrarily often.
This is essentially the frequentist notion of direct probability, \ie\ long-run frequencies under replication.
Standard tests in the Fisher and Neyman-Pearson traditions as well as Neyman confidence intervals are based on this concept of probability.
Bayesian approaches (in the now common interpretation), on the other hand, condition on the particular data and quantify inductively the probability of model parameters given the data.
The parameters are thus not fixed, and the probability is usually equated with researchers' posterior beliefs about model parameters.
There is actually a debate among Bayesians about the proper interpretation of Bayesian methods, and whether a notion of hypothesis testing is compatible (or even already contained) in the Bayesian approach.
In \citet[10]{GelmanShalizi2013}, the authors -- prominent Bayesians themselves -- acknowledge that a theory of testing is a desideratum, and they state about the standard inductive interpretation of Bayesianism that \textit{most of this received view of Bayesian inference is wrong}, and they develop a Bayesian notion of $p$ values (see also \citealp{Mayo2013}, for a frequentist reply; also \citealp{Senn2011} on different strands of Bayesianism and their stance on inductive vs.\ deductive reasoning, and \citealp{Mayo2011}, for a critical reply).
Clearly, in such quarrels between and among camps of philosophers of science and statisticians it is difficult for mere practitioners to take sides.

These quarrels relate to the second question, however.
\citet[301--302]{Divjak2016a} speaks favourably of Bayesian methods because the Bayesian concept of probability is more adequate for cognitive modeling compared to the frequentist one.
Her argument is part of a larger body of literature asking for cognitively plausible modeling techniques, for example Naive Discriminative Learning (NDL; \citealp{Baayen2011,BaayenEa2013,MilinEa2016,TheijssenEa2013}). 
On p.\ 303 of \cite{Divjak2016a}, the author goes on to explicitly mention NDL.
Yet, neither frequentist nor Bayesian methods were conceived as cognitive models, but as systems of inference for scientists (see above, and see also \citealp[302]{Divjak2016a}).
The fundamental question that lurks behind such arguments is how we interpret our statistical models (estimated on corpus data).
Are they inductive models of cognitive representations that also human learners would infer from being exposed to the corpus data?%
\footnote{In which case we are doing \textit{data science in language research} in the words of \citealp{MilinEa2016}.}
Or are they tests of theories that are pre-specified and merely tested for predictive accuracy on linguistic output data contained in corpora.
In the former case, we adopt a strong \textit{corpus as input} hypothesis and should definitely resort to methods like NDL.
However, this would most likely require us to toss most previous work done in (cognitive) linguistics into the bin and to abandon all high-level generalisations that most existing studies have been based upon.
In fact, arguments to this extreme effect and against using high-level generalisation have been made, for example in \cite{BaayenEa2016}, \citet[299--300]{Divjak2016a}, \cite{RamscarPort2016}, \cite{TheijssenEa2013}.
In the latter and less extreme case, the cognitive commitment does, however, not necessarily extend to the statistical methods used.
These methods then do not need to be any more cognitively plausible than an ANOVA used to analyse the results from an experiment.
I view my own work (see Sections~\ref{sec:germanmeasurenps}--\ref{sec:experimental}) in the tradition of testing theories (which embrace high-level generalisations), and I agree to provisionally use Likelihood methods (see below).
I remain fully agnostic with respect to the question which approach is `right'.
The best strategy for cognitive linguistics as a field might be to cultivate many methods while making sure that each method is applied carefully and competently.

@@

This concludes the general introduction.
In the remainder of the paper, I introduce the alternation phenomenon (Section~\ref{sec:descriptive}) and suggest a theory-driven set of factors influencing the alternation (Section~\ref{sec:analyses}).
Then, the corpus study is presented including an appropriate statistical analysis (Section~\ref{sec:corpusstudies}).
The two experiments which correlate results from the corpus study with experimental results are then reported (Section~\ref{sec:experimental}) before I sum up the paper in Section~\ref{sec:conclusion}.

@@

In the case at hand, three different syntactic configurations need to be distinguished \wrt case assignment inside German measure noun phrases.
If the kind noun forms an NP with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{If the kind noun is definite, the construction instantiates a true partitive.
Whereas partitives are constructions denoting a proper part-of relation as in \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities and contain indefinite kind nouns as in \textit{a sip of wine}.%
In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \cite{Hentschel1993} while some realise the difference and at least mention it \citep{Eschenbach1994,GallmannLindauer1994,Loebel1989,Zimmer2015}.
}
Here, the kind noun is in the genitive, and I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

If the kind noun is bare -- \ie\ if it does neither come with a determiner nor a modifying adjective -- it has to agree in case with the measure noun, and the genitive seen in the PGC is not acceptable, see (\ref{ex:intro:narrowapposition2}).

@@

Turning to (\ref{it:intro:plurl}), readers might have noticed that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun as in \textit{ein Sack kleine Äpfel} (\NACa) or \textit{ein Sack kleiner Äpfel} (\PGCa) `a bag of small apples', a similar alternation between PGC and NAC can be observed.
In line with experimental results reported in \citet[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (67 of 861 cases or over 92\%, cf.\ Section~\ref{sec:corpusstudies}) that the alternation cannot be analysed in the same way as in the singular case.%
\footnote{More concretely, if these cases were included in the regression analysis of the corpus data along with a factor encoding the number of the kind noun, this factor would most assuredly override any other regressor for the data points with a plural kind noun.}
While this will play a role in the interpretation of the corpus findings, MNPs with plural kind nouns will not be included in the corpus study and the experiments reported here.

@@

These construction have very idiosyncratic properties (the construction with \textit{voller} is discussed in \citealp{Zeldes2018}, the construction with \textit{mit} is discussed in \citealp{Bhatt1990}), they are not semantically equivalent to the NAC and the PGC, and they are only available for a small subset of measure nouns.
All of these constructions are only used with measure nouns denoting containers, and the whole NP always refers to the container, and never to the quantity contained in it.
They are consequently incompatible with measure nouns denoting natural portions such as \textit{Schluck} `gulp' or \textit{Haufen} `heap' and strongly grammaticalised nouns such as \textit{Gramm} `gram'.
This disqualifies them as alternatives to the \NACa\ or \PGCa\ in most contexts, and they are thus decidedly \textit{not} more ways of saying the same thing. 
It is therefore reasonable to focus on the two well-defined variants.

@@

\begin{exe}
  \ex\label{ex:determiners} 
  \begin{xlist}
    \ex{\label{ex:determiners:nac}\gll [[Drei Löffel]\Sub{Nom} [heißer Rum]\Sub{Nom}]\Sub{Nom} sind genug.\\
                                       three spoons hot rum are enough.\\
				     \glt Three spoonful of hot rum is enough.}
    \ex{\label{ex:determiners:pgc}\gll [[Einige Löffel]\Sub{Nom} [heißen Rums]\Sub{Gen}]\Sub{Nom} sind genug\\
                                       some   spoons hot    rum  are   enough.\\
				     \glt A few spoonful of hot rum is enough.}
  \end{xlist}
\end{exe}

>>>>

As for the first point, this is not quite the place to discuss it fully.
The basic distinction is a philosophical one and related to the concepts of \textit{direct} and \textit{inverse probability} (\eg\ \citealp{Senn2011}).
Frequentists assume that models and parameters are fixed, for example a model specifying that a coin is fair.
We can then calculate for observed data (for example a measurement of $3$ heads in an experiment with $10$ tosses) how often such a result or a more extreme result would occur if the model were true and we repeated the experiment arbitrarily often.
This is essentially the frequentist notion of direct probability, \ie\ long-run frequencies under replication.
Standard tests in the Fisher and Neyman-Pearson traditions as well as Neyman confidence intervals are based on this concept of probability.
Bayesian approaches (in the now common interpretation), on the other hand, condition on the particular data and quantify inductively the probability of model parameters given the data.
The parameters are thus not fixed, and the probability is usually equated with researchers' posterior beliefs about model parameters.
There is actually a debate among Bayesians about the proper interpretation of Bayesian methods, and whether a notion of hypothesis testing is compatible (or even already contained) in the Bayesian approach.
In \citet[10]{GelmanShalizi2013}, the authors -- prominent Bayesians themselves -- acknowledge that a theory of testing is a desideratum, and they state about the standard inductive interpretation of Bayesianism that \textit{most of this received view of Bayesian inference is wrong}, and they develop a Bayesian notion of $p$ values (see also \citealp{Mayo2013}, for a frequentist reply; also \citealp{Senn2011} on different strands of Bayesianism and their stance on inductive vs.\ deductive reasoning, and \citealp{Mayo2011}, for a critical reply to \citealp{Senn2011}).
Clearly, in such quarrels between and among camps of philosophers of science and statisticians it is difficult for mere practitioners to take sides.

These quarrels relate to the second point, however.
\citet[301--302]{Divjak2016a} speaks favourably of Bayesian methods because the Bayesian concept of probability is more adequate for cognitive modeling compared to the frequentist one.
Her argument is part of a larger body of literature asking for cognitively plausible modeling techniques, for example Naive Discriminative Learning (NDL; \citealp{Baayen2011,BaayenEa2013,MilinEa2016,TheijssenEa2013}). 
On p.\ 303 of \cite{Divjak2016a}, the author goes on to explicitly mention NDL as well.
Yet, neither frequentist nor Bayesian methods were conceived as cognitive models, but as systems of inference for scientists (see above, and see also \citealp[302]{Divjak2016a}).
The fundamental question that lurks behind such arguments is how we interpret our statistical models (estimated on corpus data).
Are they inductive models of cognitive representations that also human learners would infer from being exposed to the corpus data?%
\footnote{In which case we are doing \textit{data science in language research} in the words of \citealp{MilinEa2016}.}
Or are they tests of theories that are pre-specified and merely tested for predictive accuracy on linguistic output data contained in corpora.
In the former case, we adopt a strong \textit{corpus as input} hypothesis and should definitely resort to methods like NDL.
However, this would most likely require us to toss most previous work done in (cognitive) linguistics into the bin and to abandon all high-level generalisations that most existing studies have been based upon.
In fact, arguments to this extreme effect and against using high-level generalisation have been made, for example in \cite{BaayenEa2016}, \citet[299--300]{Divjak2016a}, \cite{RamscarPort2016}, \cite{TheijssenEa2013}.
In the latter and less extreme case, the cognitive commitment does, however, not necessarily extend to the statistical methods used.
These methods then do not need to be any more cognitively plausible than an ANOVA used to analyse the results from an experiment.
I view my own work (see Sections~\ref{sec:germanmeasurenps}--\ref{sec:experimental}) in the tradition of testing theories (which embrace high-level generalisations), and I agree to provisionally use Likelihood methods (see below).
I remain fully agnostic with respect to the question which approach is `right'.
The best strategy for cognitive linguistics as a field might be to cultivate many methods while making sure that each method is applied carefully and competently.

@@

This concludes the general introduction.
In the remainder of the paper, I introduce the alternation phenomenon (Section~\ref{sec:descriptive}) and suggest a theory-driven set of factors influencing the alternation (Section~\ref{sec:analyses}).
Then, the corpus study is presented including an appropriate statistical analysis (Section~\ref{sec:corpusstudies}).
Two experiments that correlate results from the corpus study with experimental results are then reported (Section~\ref{sec:experimental}) before I sum up the paper in Section~\ref{sec:conclusion}.

@@

In the case at hand, three different syntactic configurations need to be distinguished \wrt case assignment inside German measure noun phrases.
If the kind noun forms an NP with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{If the kind noun is definite, the construction instantiates a true partitive.
Whereas partitives are constructions denoting a proper part-of relation as in \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities and contain indefinite kind nouns as in \textit{a sip of wine}.
In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \cite{Hentschel1993} while some realise the difference and at least mention it \citep{Eschenbach1994,GallmannLindauer1994,Loebel1989,Zimmer2015}.
}
Here, the kind noun is in the genitive, and I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

If the kind noun is bare -- \ie\ if it comes neither with a determiner nor a modifying adjective -- it has to agree in case with the measure noun, and the genitive as seen in the PGC is not acceptable, see (\ref{ex:intro:narrowapposition2}).

@@

Turning to (\ref{it:intro:plurl}), readers might have noticed that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun as in \textit{ein Sack kleine Äpfel} (\NACa) or \textit{ein Sack kleiner Äpfel} (\PGCa) `a bag of small apples', a similar alternation between PGC and NAC can be observed.
In line with experimental results reported in \citet[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (67 of 861 cases or over 92\%, cf.\ Section~\ref{sec:corpusstudies}) that the alternation cannot be analysed in the same way as in the singular.%
\footnote{More concretely, if these cases were included in the regression analysis of the corpus data along with a factor encoding the number of the kind noun, this factor would most assuredly override any other regressor for the data points with a plural kind noun.}
While this will play a role in the interpretation of the corpus findings, MNPs with plural kind nouns will not be included in the corpus study and the experiments reported in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.

@@

These construction have very idiosyncratic properties (the construction with \textit{voller} is discussed in \citealp{Zeldes2018}, the construction with \textit{mit} is discussed in \citealp{Bhatt1990}), they are not semantically equivalent to the NAC and the PGC, and they are only available for a small subset of measure nouns.
All of these constructions are used only with measure nouns denoting containers, and the whole NP always refers to the container, and never to the quantity contained in it.
They are consequently incompatible with measure nouns denoting natural portions such as \textit{Schluck} `gulp' or \textit{Haufen} `heap' and strongly grammaticalised nouns such as \textit{Gramm} `gram'.
This disqualifies them as alternatives to the \NACa\ or \PGCa\ in most contexts, and they are thus decidedly \textit{not} more ways of saying the same thing. 
It is therefore reasonable to focus on the two well-defined variants.

@@

\begin{exe}
  \ex\label{ex:determiners} 
  \begin{xlist}
    \ex{\label{ex:determiners:nac}\gll [[Drei Löffel]\Sub{Nom} [heißer Rum]\Sub{Nom}]\Sub{Nom} sind genug.\\
                                       three spoons hot rum are enough\\
				     \glt Three spoonful of hot rum is enough.}
    \ex{\label{ex:determiners:pgc}\gll [[Einige Löffel]\Sub{Nom} [heißen Rums]\Sub{Gen}]\Sub{Nom} sind genug.\\
                                       some   spoons hot    rum  are   enough\\
				     \glt A few spoonful of hot rum is enough.}
  \end{xlist}
\end{exe}
